---
output:
  word_document: default
  html_document: default
---

```markdown
---
title: "Sales Analysis and Forecasting"
author: "Mark Abong'o"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installing and Loading Packages

```{r packages, message=FALSE, warning=FALSE}
# install.packages("forecast") # Uncomment if not already installed

library(ggplot2)
library(tidyverse)
library(readr)
library(forecast)
```

## Loading Sales Data

```{r load-data}
sales_data <- read_csv("D:/Documents/train.csv")  # Make sure path is correct
```

## Data Exploration

```{r explore}
dim(sales_data)
head(sales_data)
names(sales_data)
glimpse(sales_data)
```

## 🧹 Data Cleaning

```{r clean-data}
missingValues <- !complete.cases(sales_data)
sales_data[missingValues,]
sales_data <- na.omit(sales_data)

# Confirm missing row (2235) is gone
sales_data[2235, 1]
```

## Date Conversion

```{r date-formatting}
date_columns <- grep("Date", names(sales_data), value = TRUE, ignore.case = TRUE)
date_columns

sales_data$`Order Date` <- as.Date(sales_data$`Order Date`, format = "%d/%m/%Y")
sales_data$`Ship Date` <- as.Date(sales_data$`Ship Date`, format = "%d/%m/%Y")

glimpse(sales_data)
```

## Exploratory Data Analysis

### Summary of Sales

```{r summary}
summary(sales_data$Sales)
```

### Boxplot: Sales per Product Category

```{r boxplot-category}
ggplot(sales_data, aes(x = Category, y = Sales)) +
  geom_boxplot(fill = "steelblue", width = 0.8, notch = TRUE, outlier.colour = "red", outlier.shape = 8) +
  labs(title = "Sales per Product Category", x = "Categories", y = "Sales") +
  theme_minimal()
```

### Line Chart: Total Sales Over Time

```{r sales-over-time}
ggplot(sales_data, aes(x = `Order Date`, y = Sales)) +
  geom_line(color = "steelblue") +
  labs(title = "Total Sales Over Time", x = "Date", y = "Sales")
```

### Bar Chart: Top Selling Categories

```{r top-categories}
ggplot(sales_data, aes(x = Category, y = Sales, fill = Category)) +
  geom_bar(stat = "identity", width = 0.5) +
  scale_fill_manual(values = c("red", "blue", "green")) +
  labs(title = "Top Selling Categories", x = "Categories", y = "Sales") +
  theme_minimal()
```

### Bar Chart: Sales by Region

```{r sales-by-region}
ggplot(sales_data, aes(x = Region, y = Sales, fill = Region)) +
  geom_bar(stat = "identity", width = 0.5) +
  labs(title = "Sales by Region", x = "Regions", y = "Sales") +
  coord_flip() +
  theme_minimal()
```

### Line Chart: Customer Purchase Frequency

```{r customer-frequency}
ggplot(sales_data, aes(x = `Customer ID`, y = Sales)) +
  geom_line(color = "steelblue") +
  labs(title = "Customer Purchase Frequency", x = "Customer ID", y = "Sales") +
  theme_classic()
```

## Forecasting with ARIMA

```{r forecasting}
sales_ts <- ts(sales_data$Sales, start = c(2020, 1), frequency = 12)
arima_model <- auto.arima(sales_ts)
forecasted_sales <- forecast(arima_model, h = 12)

autoplot(forecasted_sales) + labs(title = "Sales Forecast for Next 12 Months")
```

## Export Cleaned Data

```{r save-data}
write_csv(sales_data, "cleaned_customer_sales.csv")
```
```