---
title: "Data Exploration and Visualization with Gapminder"
author: "Mark"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

# Introduction
This project explores the `gapminder` dataset using **R** and **tidyverse**. We perform **data cleaning, descriptive analysis, and visualizations**, along with statistical tests such as **t-tests** and **ANOVA**.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Installing and Loading Packages
```{r install-packages, eval=FALSE}
install.packages("rmarkdown")
install.packages("shiny")
```

```{r load-packages}
library(ggplot2)
library(tidyverse)
library(gapminder)
```

# Data Exploration
```{r data-exploration}
head(gapminder)
tail(gapminder)
glimpse(gapminder)
summary(gapminder)
unique(gapminder$country)
class(gapminder$pop)
```

# Data Cleaning
```{r data-cleaning}
noMissingData <- drop_na(gapminder)
missingData <- !complete.cases(gapminder)
gapminder[missingData,]

cleaned_data <- gapminder %>% 
  select(country, pop, year) %>% 
  rename(population = pop) %>% 
  filter(year > 2000 & population > 10000000) %>% 
  arrange(year)
head(cleaned_data)
```

```{r reshape-data}
sampleData <- select(gapminder, country, pop, year)
yearly_data <- sampleData %>% 
  pivot_wider(names_from = year, values_from = pop)
View(yearly_data)
```

# Descriptive Statistics
```{r descriptive-statistics}
population <- gapminder$pop
summary(population)
population <- population / 1000000
summary(population) 
```

# Data Visualization
## Bar Chart of Continents
```{r bar-chart}
ggplot(data=gapminder, aes(x=continent)) + 
  geom_bar(fill = "steelblue") + 
  labs(title = "Continents", x="Habitable Continents")
```

## Scatter Plot of Population (African Countries)
```{r scatter-plot}
gapminder %>% 
  filter(year > 2006 & continent == "Africa" & pop > 40000000) %>% 
  ggplot(aes(x = country, y = pop/1000000)) + 
  geom_point(size=5, alpha = 0.5) +
  labs(title = "Population of African Countries", y="Population (millions)")
```

# Statistical Analysis
## T-test (Life Expectancy in Americas vs Oceania)
```{r t-test}
gapminder %>% 
  filter(continent %in% c("Americas", "Oceania")) %>%  
  t.test(lifeExp ~ continent, data = .)
```

## ANOVA (Life Expectancy Across Continents)
```{r anova}
gapminder %>% 
  filter(year == 2007, continent %in% c("Asia", "Africa", "Americas")) %>% 
  aov(lifeExp ~ continent, data = .) %>%
  summary()
```

# Conclusion
This project explored the `gapminder` dataset, cleaned the data, visualized key insights, and performed statistical tests. The findings suggest patterns in population distribution and life expectancy differences among continents.

